---
title: "Updating Data for Portal Apps"
author: "Brian Waismeyer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Updating Data for Portal Apps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Data Portal brings together a large family of applications that allow users
to explore and visualize POC managed data.

The Data Portal team requires that all applications have their data source and
generation process documented and - ideally - automated. With rare exceptions
(e.g., the COS application), all applications should receive data ready for
consumption either directly from a POC SQL database or from an R supported
update process.

`pocr` has a number of functions that support updating application data for
those applications that are dependent on R data processing. These functions
are tied together by `get_portal_app_data()` - a wrapper that calls the R 
processes needed to retrieve and format data for R-dependent Data Portal
applications.

This vignette will walk the user through the process of updating R-dependent
Data Portal applications, including using the supporting `pocr` functions.

## Install Supporting Tools
Prior to starting the update process, you need to prepare the following tools.

1. [GitHub account](https://github.com/)
    
    First things first, you need a GitHub account, you need to be part of
    [pocdata](https://github.com/pocdata), and you need permissions to 
    view/access any apps you intend to update. Membership in the "Data Portal"
    team should get you the permissions.
    
2. [Git](https://git-scm.com/downloads)
    
    Now you need a way to receive files from and pass files to GitHub. Install
    Git (specifically we want Git Bash) using all the default settings.
    
3. [R](https://cran.r-project.org/)
    
    This is an R vignette, so I'm guessing you already did this. But just in
    case you don't have R... you need it to use the `pocr` package.
    
4. [RStudio](https://www.rstudio.com/products/RStudio/#Desktop)
    
    Not required, but makes working with R much more pleasant. You want the 
    free desktop version.

5. `devtools`
    
    `devtools` is an R package that supports building packages. We need it
    to install and build the `pocr` package.
    
    Open RStudio or an R console and install `devtools` from the R command
    line.
    
    ```{r, eval = FALSE}
    install.packages("devtools", depends = TRUE)
    ```
    
6. ['pocr'](https://github.com/pocdata/pocr)
    
    `pocr` is the POC R package with the data retrieval/formatting functions,
    along with a variety of other useful functions for POC data work.
    
    Open RStudio or an R console and install `pocr` from the R command line.
    ```{r, eval = FALSE}
    devtools::install_github("pocdata/pocr")
    ```

7. Connections to the POC SQL server and the annie MySQL server
    
    Talk to Gregor or your senior colleagues for how to set these up. Record
    the names you give the odbc connections. I suggest using "POC" for the POC
    SQL server and "annie" for the annie MySQL server.

Once you've installed the above materials, you will need to 
[set up Git](https://help.github.com/articles/set-up-git/#setting-up-git) with
your credentials.

Depending on your permissions and platform, you may also need to configure
your R library path. This process varies by operating system. Pester a
fellow Data Portal team member and/or Google to setup your path.

## Clone the Repos You Want to Update
Git allows us to create local, linked copies (aka - "clones") of the 
repositories on GitHub. We can makes changes to these copies - such as
changing the data files - and then "push" our changes to the GitHub
repository.

Once you have your update tools installed and configured, you will want
to create clones of all the repos you want to update.

For a list of applications - and their repo names - that are currently
supported by `pocr`, please see the Data Portal
[application portfolio](https://docs.google.com/spreadsheets/d/1qrLbi2bCfnYeVRNlf89QYhMr1piQ0eDFGVRSZQz1ksM/edit?usp=sharing).

I recommend that you create all your clones in a common location (e.g.,
`C:/Projects/`) so that they are easy to locate and work with

To clone a repo:

* open Git Bash
* `cd` to your target directory (e.g., `cd C:/Projects`)
* type and enter `git clone repo-url` (e.g., 
    `git clone https://github.com/pocdata/pocr`)
* inspect the contents of the target folder with File Explore or `ls` and 
    you should see a copy of the targe repo (e.g., `C:/Projects/pocr/`)

## Verify that Upstream Data Dependencies are Up-to-Date
**TBD**

You can ignore this step for now. We are still resolving where in the process
this should occur and what this check should entail.

## Run `get_portal_app_data()` to Retrieve/Format Current Data
At this point, you need to generate new data files for the applications you
are aiming to update.

Open an R console or RStudio, load `pocr`, and execute `get_portal_app_data()`.

```{r, eval = FALSE}
library(pocr)

get_portal_app_data()
```    

You may need to adjust the arguments to `get_portal_app_data()` to match
the names you gave to the POC and annie odbc connections and/or to match
the specific applications you want to update.

By default:

* all (R dependent) applicatiosn will have data retrieved for them
* poc_connection = "POC"
* annie_connection = "annie"

Observe the update process - `get_portal_app_data()` will report what apps
it is trying to retrieve/format data for and will report if each appears to
succeed.

If you observe any errors, you will need to inspect the errors to try and
determine the cause of the issue. This may require that you look up the 
helper function associated with a given application (e.g., 
`get_county_dashboard_data()` is the function called by `get_portal_app_data()` 
to update the County Dashboard app).

## Manually Inspect the Data for Obvious Errors
`get_portal_app_data()` will create a folder called `app_data` (or a numbered
variant, such as `app_data1`) each time it is run. This folder will have
a subfolder for each target app repo (e.g., `app_data/county_dashboard`).

* Check that the `app_data` folder has been created and subfolders have been
    made for all targeted apps.
* Then open the subfolders and check that they have files in them and that 
    the files are in the correct format for the target application. Peek inside
    your copy of the target app repo if you are unsure what format is correct.
* Open at least a few of the data files and make sure they have content and
    that this content seems apprpriate for the target app. Again, you may
    want to compare against the target app files directly.

## Copy the New Datafiles to Your Copy of the Target App Repo
For each app you want to update, you need to replace the data files in
your local copy of the repo with the new data files.

As an example, updating `county_dashboard` places a bundle of .csv files
in `app_data/county_dashboard`. To update your local copy of the County
Dashboard repo, you would:

* navigate the your local repo data directory 
    (e.g., `C:/Projects/county_dashboard/data`)
* delete the current data files
* replace these with the new data files from `app_data/county_dashboard`

## Update the Target App Repo on GitHub
Once you have updated your local copies of the target app repos, you will
need to push your changes to GitHub.

* open Git Bash
* navigate to the directory of your local copy (e.g., 
`cd C:/Projects/county_dashboard`)
* tell Git to track all your changes to the repo (`git add -A`)
* tell Git to "commit" the changes for a push and provide a concise update
    message (`git commit -m "8-15-2015 data update"`)
* tell Git to "push" the changes to GitHub (`git push`)
* you may be prompted for your GitHub credentials - provide them and then
    watch to make sure the push completes

Complete the above process for all the apps you have generated new data for
and VOILA you're just about done!

## Inform the Site Manager What Apps Have Updated Data
Contact the Data Portal site manager (Erika Deal: edeal@uw.edu) and let her
know which apps have updated data. She will complete the process to move the
data to the live apps.
